Deciding Bitvector Termination with Program Synthesis

### 2014-07-02 papers/termination/kalashnikov.tex

We consider the following parameters:

\subsubsection{Program Length: $l$}
The first parameter we introduce is program length, denoted by $l$.
At each iteration we synthesise programs of length exactly $l$.
We start with $l = 1$ and increment $l$ whenever we determine
that no program of length $l$ can satisfy the specification.  When we do
successfully synthesise a program, we are \emph{guaranteed that it
is of minimal length} since we have previously established that no
shorter program is correct.

\subsubsection{Word Width: $w$}
An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
has its own set of parameters.  The only relevant parameter is
the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
in each internal register and immediate constant.  This parameter is denoted by
$w$.  The size of the final SAT problem generated by {\sc cbmc} scales
polynomially with $w$, since each intermediate C variable corresponds
to $w$ propositional variables.

It is often the case that a program which satisfies the specification
on an $\mathcal{L}$-machine with $w = k$ will continue to satisfy the
specification when run on a machine with $w > k$.  For example, the program
in Fig.~\ref{fig:bitvector-program} isolates the least-significant bit of a word.
This is true irrespective of the word size of the machine it is run on -- it will
isolate the least-significant bit of an 8-bit word just as well as it will a
32-bit word.  An often successful strategy is to synthesise a program for an
$\mathcal{L}$-machine with a small word size and then to check whether the
same program is correct when run on an $\mathcal{L}$-machine with a
full-sized word.

The only wrinkle here is that we will sometimes synthesise a program containing
constants.  If we have synthesised a program with $w=k$,
the constants in the program will be $k$-bits wide.  To extend the program
to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
numbers from $k$-bit ones.  We have several strategies for this and
just try each in turn.  Our strategies are shown in Fig.~\ref{fig:generalize}.
$\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
and $b \cdotp c$ means the concatenation of bitvectors $b$ and $c$.

\begin{figure}

\centering
\begin{minipage}[t]{.45\textwidth}
\begin{eqnarray*}
 \bv{m}{m} & \rightarrow & \bv{n}{n} \\
 \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
 \bv{m+1}{m} & \rightarrow & \bv{n+1}{n}
\end{eqnarray*}
\end{minipage}
\begin{minipage}[t]{.45\textwidth}
\begin{eqnarray*}
 \bv{x}{m} & \rightarrow & \bv{x}{n} \\
 \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
 \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
\end{eqnarray*}
\end{minipage}

\caption{Rules for extending
an $m$-bit wide number to an $n$-bit wide one.
 \label{fig:generalize}}
\end{figure}

Sometimes a program will be correct for some particular word width $w$, but is
not correct for $w' > w$ even if the constants are replaced with appropriate ones.
When we detect this situation, we increase $w$ and continue synthesising.

\subsubsection{Number of Constants: $c$}
Instructions in $\mathcal{L}$ take either one or two operands.
Since any instruction whose operands are all constants can always be
eliminated (since its result is a constant), we know that a loop-free program
of minimal length will not contain any instructions with two constant
operands.  Therefore the number of constants that can appear in
a minimal program of length $l$ is at most $l$.  By minimising the number
of constants appearing in a program, we are able to use a particularly
efficient program encoding that speeds up the synthesis procedure
substantially.  The number of constants used in a program is the parameter $c$.

$\mathcal{L}$ is an SSA, three-address instruction set\footnote{
We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
the programs to be smaller and synthesis to be faster as a result.  We saw
the opposite effect -- the more complex interpreter led to much slower synthesis.
}.  Destination registers
are implicit and a fresh register exists for each instruction to write its
output to.  A na\"{\i}ve way to encode $\mathcal{L}$ instructions is to have an
opcode and two operands, where each operand is either a register (i.e., a program argument
or the result of a previous instruction), or an immediate constant.

In this encoding, each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
of instruction types in $\mathcal{L}$.  Each operand can be encoded using
$\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word width, plus one
bit to specify whether the operand is a register name or an immediate constant.
One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
For an $n$-instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
the entire program.

If we instead limit the number of constants that can appear in the program,
our operands can be encoded using fewer bits.  For an $n$-instruction program
using $c$ constants and taking $a$ arguments as inputs, each operand can refer
to a program argument, the result of a previous instruction or a constant.
This can be encoded using $\lceil \log_2 (c+a+n-1) \rceil$ bits, which means each instruction
can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (c + a + n - 1) \rceil$ and the full program
needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (c + a + n - 1) \rceil + cw$$ bits to encode.

We give an example. Our language $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
For a 10-instruction program over 1 argument, using 2 constants on a 32-bit word
machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
Using the second encoding, each operand can be represented using
$\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
This is a substantial reduction in size and when the desired program requires
only few constants this can lead to a very significant speed up.

As with program length, we progressively increase the number of constants in
our program.  We start by trying to synthesise a program with no constants,
then if that fails we attempt to synthesise using one constant and so on until we
reach $c = l$.

\subsubsection{Searching the Program Space}

>>>>

We consider the following parameters:
\begin{itemize}
\item{Program Length: $l$}
%% The first parameter we introduce is program length, denoted by $l$.
%% At each iteration we synthesise programs of length exactly $l$.
%% We start with $l = 1$ and increment $l$ whenever we determine
%% that no program of length $l$ can satisfy the specification.  When we do
%% successfully synthesise a program, we are \emph{guaranteed that it
%% is of minimal length} since we have previously established that no
%% shorter program is correct.
\item{Word Width: $w$}
%% An $\mathcal{L}$-program runs on a virtual machine (the $\mathcal{L}$-machine) that
%% has its own set of parameters.  The only relevant parameter is
%% the \emph{word width} of the $\mathcal{L}$-machine, that is, the number of bits
%% in each internal register and immediate constant.  This parameter is denoted by
%% $w$.  The size of the final SAT problem generated by {\sc cbmc} scales
%% polynomially with $w$, since each intermediate C variable corresponds
%% to $w$ propositional variables.

%% It is often the case that a program which satisfies the specification
%% on an $\mathcal{L}$-machine with $w = k$ will continue to satisfy the
%% specification when run on a machine with $w > k$.  For example, the program
%% in Fig.~\ref{fig:bitvector-program} isolates the least-significant bit of a word.
%% This is true irrespective of the word size of the machine it is run on -- it will
%% isolate the least-significant bit of an 8-bit word just as well as it will a
%% 32-bit word.  An often successful strategy is to synthesise a program for an
%% $\mathcal{L}$-machine with a small word size and then to check whether the
%% same program is correct when run on an $\mathcal{L}$-machine with a
%% full-sized word.

%% The only wrinkle here is that we will sometimes synthesise a program containing
%% constants.  If we have synthesised a program with $w=k$,
%% the constants in the program will be $k$-bits wide.  To extend the program
%% to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
%% numbers from $k$-bit ones.  We have several strategies for this and
%% just try each in turn.  Our strategies are shown in Fig.~\ref{fig:generalize}.
%% $\mathcal{BV}(v, n)$ denotes an $n$-bit wide bitvector holding the value $v$
%% and $b \cdotp c$ means the concatenation of bitvectors $b$ and $c$.

%% \begin{figure}

%% \centering
%% \begin{minipage}[t]{.45\textwidth}
%% \begin{eqnarray*}
%%  \bv{m}{m} & \rightarrow & \bv{n}{n} \\
%%  \bv{m-1}{m} & \rightarrow & \bv{n-1}{n} \\
%%  \bv{m+1}{m} & \rightarrow & \bv{n+1}{n}
%% \end{eqnarray*}
%% \end{minipage}
%% \begin{minipage}[t]{.45\textwidth}
%% \begin{eqnarray*}
%%  \bv{x}{m} & \rightarrow & \bv{x}{n} \\
%%  \bv{x}{m} & \rightarrow & \bv{x}{m} \cdotp \bv{0}{n - m} \\
%%  \bv{x}{m} & \rightarrow & \underbrace{\bv{x}{m} \cdotp \ldots \cdotp \bv{x}{m}}_{\frac{n}{m} \mathrm{ times}}
%% \end{eqnarray*}
%% \end{minipage}

%% \caption{Rules for extending
%% an $m$-bit wide number to an $n$-bit wide one.
%%  \label{fig:generalize}}
%% \end{figure}

%% Sometimes a program will be correct for some particular word width $w$, but is
%% not correct for $w' > w$ even if the constants are replaced with appropriate ones.
%% When we detect this situation, we increase $w$ and continue synthesising.

\item{Number of Constants: $c$}
%% Instructions in $\mathcal{L}$ take either one or two operands.
%% Since any instruction whose operands are all constants can always be
%% eliminated (since its result is a constant), we know that a loop-free program
%% of minimal length will not contain any instructions with two constant
%% operands.  Therefore the number of constants that can appear in
%% a minimal program of length $l$ is at most $l$.  By minimising the number
%% of constants appearing in a program, we are able to use a particularly
%% efficient program encoding that speeds up the synthesis procedure
%% substantially.  The number of constants used in a program is the parameter $c$.

%% $\mathcal{L}$ is an SSA, three-address instruction set\footnote{
%% We experimented with implementing $\mathcal{L}$ as a stack machine, expecting
%% the programs to be smaller and synthesis to be faster as a result.  We saw
%% the opposite effect -- the more complex interpreter led to much slower synthesis.
%% }.  Destination registers
%% are implicit and a fresh register exists for each instruction to write its
%% output to.  A na\"{\i}ve way to encode $\mathcal{L}$ instructions is to have an
%% opcode and two operands, where each operand is either a register (i.e., a program argument
%% or the result of a previous instruction), or an immediate constant.

%% In this encoding, each opcode requires $\lceil \log_2 I \rceil$ bits to encode, where $I$ is the number
%% of instruction types in $\mathcal{L}$.  Each operand can be encoded using
%% $\log_2 w$ bits, where $w$ is the $\mathcal{L}$-machine word width, plus one
%% bit to specify whether the operand is a register name or an immediate constant.
%% One instruction can therefore be encoded using $\lceil \log_2 I \rceil + 2w + 2$ bits.
%% For an $n$-instruction program, we need $$\lceil n \log_2 I \rceil + 2nw + 2n$$ bits to encode
%% the entire program.

%% If we instead limit the number of constants that can appear in the program,
%% our operands can be encoded using fewer bits.  For an $n$-instruction program
%% using $c$ constants and taking $a$ arguments as inputs, each operand can refer
%% to a program argument, the result of a previous instruction or a constant.
%% This can be encoded using $\lceil \log_2 (c+a+n-1) \rceil$ bits, which means each instruction
%% can be encoded in $\lceil \log_2 I \rceil + \lceil \log_2 (c + a + n - 1) \rceil$ and the full program
%% needs $$\lceil n \log_2 I \rceil + \lceil n \log_2 (c + a + n - 1) \rceil + cw$$ bits to encode.

%% We give an example. Our language $\mathcal{L}$ has 15 instruction types, so each opcode is 4 bits.
%% For a 10-instruction program over 1 argument, using 2 constants on a 32-bit word
%% machine the first encoding requires $10 * (4 + 32 + 1 + 32 + 1) = 700$ bits.
%% Using the second encoding, each operand can be represented using
%% $\log_2 (2 + 1 + 10 - 1) = 4$ bits, and the entire program requires 184 bits.
%% This is a substantial reduction in size and when the desired program requires
%% only few constants this can lead to a very significant speed up.

%% As with program length, we progressively increase the number of constants in
%% our program.  We start by trying to synthesise a program with no constants,
%% then if that fails we attempt to synthesise using one constant and so on until we
%% reach $c = l$.

\end{itemize}
\subsubsection{Searching the Program Space}
